---
title: "DiagTestKit Package"
subtitle: |
  | Getting Started
author: Center for Veterinary Biologics - Statistics Section
date: December 2017
output: pdf_document
toc: true
number_sections: true
numbersections: true
fontsize: 12pt
urlcolor: blue
vignette: >
  %\VignetteIndexEntry{DiagTestKit Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, tidy = FALSE)
options(width = 50)

```

# Introduction

The `DiagTestKit` package provides functions to obtain point and interval estimates for diagnostic sensitivity and specificity for an experimental diagnostic test kit when one or more imperfect reference tests are used.  The package also includes a function to estimate and obtain confidence intervals for diagnostic sensitivity or specificity in the event an infallible reference test is used.

The technical details for how the algorithm works are provided in [STATWI0002](https://www.aphis.usda.gov/aphis/ourfocus/animalhealth/veterinary-biologics/biologics-regulations-and-guidance/ct_vb_statwi).  For a 2-state experimental test with one or more fallible reference tests, the default display will show the point and interval estimates for sensitivity ($\pi_1$) and specificity ($\theta_1$).  For a 3-state experimental test with one or more fallible reference tests, the default display will show the point and interval estimates for sensitivity ($\pi_1$), specificity ($\theta_1$), probability of suspect for disease positive ($\psi_1$) and probability of suspect for disease negative ($\phi_1$).  The algorithm is optimizing the sum of squared residuals with respect to $\pi_1$, $\theta_1$, $\delta_1$, and $\gamma_1$ where $\delta_1$ and $\gamma_1$ express the probability of a suspect result as a fraction of the non-correct test results.  See [STATWI0002](https://www.aphis.usda.gov/aphis/ourfocus/animalhealth/veterinary-biologics/biologics-regulations-and-guidance/ct_vb_statwi) for further details.

This vignette provides step--by--step instructions to using the package for obtaining point and interval estimates for the sensitivity (Sn = $\pi$) and specificity (Sp = $\theta$) of a 2--state experimental test kit. To run the `DiagTestKit` package requires a functional installation of R software.  The `DiagTestKit` package is available on (need to include a hyperlink to gitHub)

# Data processing

The data file titled dichotomoussensspec_deviceinfo.csv will be used to illustrate the use of the `estimateSnSp` function.  While this data file includes 3 sample types, only the whole blood samples will be used to demonstrate the `estimateSnSp` function.  Importing the data is illustrated for completeness, but the dataset dat_dichot is available within the `DiagTestKit` package.  All raw data files should follow the formats in Section 1.8 or Section 1.9 of the [CVB Data Guide](https://www.aphis.usda.gov/animal_health/vet_biologics/publications/16-CVBDataGuide.pdf). A table of counts (dat_infal) available within the 'DiagTestKit' package will be used to illustrate the use of the `cloppearSnSp` function.  The `cloppearSnSp` provides the binomail confidence interval using the Clopper-Pearson method when an infallible reference test is used.  

## Load package and data
```{r,echo=TRUE, warning=FALSE, message=FALSE}
library(ggplot2)
library(DiagTestKit)
data("ExampleData")
dat <- dat_dichot
infallible <- dat_infal
```

## Obtaining counts

The data input for the `estimateSnSp` function requires creating a data frame that includes the count for each unique combination of test results.  The data frame should include zero counts for possible test combinations that were not observed, but it should not include zero counts for impossible test combinations (structural zeros).  Including non-structural zeros may be achieved using the .drop=FALSE setting within `ddply`.  There is more than one way to obtain the data frame with the counts for all possible test combinations.  Two possibilities are illustrated here.   First, the `ddply` function in the `plyr` package can be used. 
```{r,echo=TRUE,eval=T}
#Blood Samples
blood <- ddply(.data = subset(dat, specimen == 'wholeblood'),
                  .variables = .(visual_read, ref_result),
                  .drop = FALSE,
                  .fun = summarize,
                  Count = length(deviceID))
```





This will require renaming the columns. The columns need to be named in a manner that will be recognized by `estimateSnSp`.  Namely, the column name of visual_read from the raw data file needs to be changed to reflect this is the response associated with the experimental kit.  The expression 'exp' should be included in the column corresponding to the experimental test results.  The expression 'ref' should be included in the all columns corresponding to the reference test(s) results. 

 
```{r, echo=T,eval=T} 
names(blood) <- c('Exp', 'Ref1', 'Count')
```

Alternatively, the data frame including the counts from all possible testing combinations can be obtained by converting the output from the `table` function in `base` R into a data frame.


```{r, echo=T,eval=T}
blood2 <- data.frame(table(Exp = dat$visual_read[dat$specimen == 'wholeblood'],
                          Ref1 = dat$ref_result[dat$specimen == 'wholeblood']))
```

# Fallible Reference Test(s)

Use the `estimateSnSp` function to estimate the sensitivity and specificity of the 2--state experimental test.  The function has 4 required inputs, the data frame of counts (dat), reasonable values based on available information for the sensitivity of each reference test (Sn.ref), reasonable values based on available information for the specificity of each reference test (Sp.ref), and a named vector containing reasonable estimates of prevalence for each population sampled.  Two alternative methods for inputting the performance characteristics (sensitivity and specificity) of the reference test are available (and illustrated below) for a 2--stage reference test, a named vector or a named data frame.  For a 3--stage reference test, the probability of suspect as a proportion of an incorrect test result (for both disease positive and disease negative) must be supplied.  Therefore, if one or more reference test(s) have 3 stages, a named data frame must be used.


## Use a named data frame for inputs of Sn.ref and Sp.ref
This example illustrates the use of a named data frame rather than a named vector for the input variables Sn.ref and Sp.ref.  The output seen here indicates that there was no column named 'population' and the data set is being treated as if it has been sampled from a single population.  The function also provides updates for every 50 iterations (simulation cycles).  Here nsim was not specified in the input and therefore the default of 1000 simulation cycles is being performed.

```{r,echo=TRUE,eval=T,results="hide"}
blood_SnSp <- estimateSnSp(dat = blood,
                           Sn.ref = data.frame(Ref1 = c(0.95, 0)),
                           Sp.ref = data.frame(Ref1 = c(0.98, 0)),
                           prev.pop = (A = 0.70))
```

`## The optimization has begun `  
`## The following is the number of iterations completed: 50`  
`## The following is the number of iterations completed: 100`  
`## The following is the number of iterations completed: 150`  
$\vdots$  
`## The following is the number of iterations completed: 1000`  
  
```{r,echo=T,eval=T}
blood_SnSp

```
## Use a named vector for inputs of Sn.ref and Sp.ref

As this example has a single 2--state reference test, a named vector can be used for the input variables for Sn.ref and Sp.ref (rather than a named data frame).  If one or more of the reference tests are 3--state tests (includes a suspect region), using a named vector is not an option.  Using named vectors for Sn.ref and Sp.ref is illustrated with the data frame obtained by converting the output from the `table` function (blood2).  This example will also reduce the frequency of the message regarding the number of iterations complete.

```{r,echo=TRUE,eval=T}
blood2_SnSp <- estimateSnSp(dat = blood2,
                           Sn.ref = c(Ref1 = 0.95),
                           Sp.ref = c(Ref1 = 0.98),
                           prev.pop = (A = 0.70),
                           control = estimateSnSpControl(iter.n = 250))
blood2_SnSp
```


## Controlling Output

### Seeds
The output from the 2 examples is not exactly the same because the seed was not set in these examples, but the seed used can be viewed.

```{r, echo=T,eval=T}
blood_SnSp$input$seed
blood2_SnSp$input$seed
```

The output can be forced to be the same by controlling the seed.  This will be illustrated by using the second data set and the seed used for the first call to `estimateSnSp`.  The message regarding the number of iterations has been suppressed in this example.

```{r, echo=T, eval=T}
blood2_a_SnSp <- 
  estimateSnSp(dat = blood2,
      Sn.ref = c(Ref1 = 0.95),
      Sp.ref = c(Ref1 = 0.98),
      prev.pop = c(A = 0.70),
      control = estimateSnSpControl(seed = blood_SnSp$input$seed,
                                      rep.iter = FALSE))
blood2_a_SnSp
```


## Evaluating Output

### Convergence

Prior to reporting the point and interval estimates, verify the optimization convergence in each instance.  A numeric value of 0 indicates
the optimization converged.  Here, we see that all have converged and we can check the convergence message as well.

```{r, echo=T,eval=T}
unique(blood_SnSp$detailOut$Converge)
unique(blood_SnSp$detailOut$Message)
```

### Distributions of Optimized Values

All optimized values for sensitivity can be viewed using a histogram.

```{r, echo=F,eval=T}
Sn.fig <- ggplot() +
          geom_histogram(aes(blood_SnSp$detailOut$Exp.Sn), binwidth = 0.005) +
          labs(x = 'Experimental Test Sensitivity Estimates', y = 'Count')
Sn.fig

```
\pagebreak

All optimized values for specificity can be viewed using a histogram.

```{r,echo=F,eval=T}
Sp.fig <- ggplot() +
          geom_histogram(aes(blood_SnSp$detailOut$Exp.Sp), binwidth = 0.005)+
          labs(x = 'Experimental Test Specificity Estimates', y = 'Count')
Sp.fig
```
\pagebreak


### Simulated Values for Reference Test
Similarly, the simulated values of sensitivity for the reference test can be extracted using `blood_SnSp$input$Sn.sims`.  The simulated
distribution for the sensitivity of the reference test is displayed as a histrogram.

```{r, echo=F,eval=T}
Sn.ref <- ggplot()+
          geom_histogram(aes(blood_SnSp$input$Sn.sim[, 1]), binwidth = 0.025)+
          labs(x = 'Simulated Sensitivity for Reference', y = 'Count')
Sn.ref
```

\pagebreak

The simulated values of specificity for the reference test can be extracted using `blood_SnSp$input$Sp.sims`.  The simulated
distribution for the specificity of the reference is displated as a histrogram.

```{r, echo=F,eval=T}
Sp.ref <- ggplot()+
          geom_histogram(aes(blood_SnSp$input$Sp.sim[, 1]), binwidth = 0.05)+
          labs(x = 'Simulated Specificity for Reference', y = 'Count')
Sp.ref
```

#Infallible Reference Test

Use the `cloppearSnSp` function to estimate the sensitivity and specificity of a 2--state experimental test when an infallible reference test has been used to 
determine the true disease status of each sample.  The function has one required input, the data frame of counts (dat).    

## Estimating Sensitivity
By default the function will estimate sensitivity.  

```{r, echo=T,eval=T}
infal_Sn <- cloppearSnSp(dat = infallible)
infal_Sn
```
The total number of positive samples tested can be extracted using `infal_Sn$data$Total.Positive` and the number of positive samples that tested positive by the experimental test can be obtained
using `infal_Sn$data$Test.Positive`.  The sensitivity estimate can be obtained by `infal_Sn$calcVal$Sn` and the lower and upper confidence limits can be obtained using `infal_Sn$calcVal$Sn.LL` and 
`infal_Sn$calcVal$Sn.UL`, respectively.

## Estimating Specificity
The function will estimate specificity when `est.Sn = FALSE`.

```{r, echo = T, eval = T}
infal_Sp <- cloppearSnSp(dat = infallible, 
                         est.Sn = FALSE)
infal_Sp
```

Similarly, the total number of negative samples tested can be extracted using `infal_Sp$data$Total.Negative` and the number of negative samples that tested negative by the experimental test can be obtained
using `infal_Sp$data$Test.Negative`.  The specificity estimate can be obtained by `infal_Sp$calcVal$Sp` and the lower and upper confidence limits can be obtained using `infal_Sp$calcVal$Sp.LL` and 
`infal_Sp$calcVal$Sp.UL`, respectively.











